---
title: Conditional Means
author: LPO 9952 | Spring 2016
output: md_document
---


```{r, echo = FALSE, message = FALSE}
require(knitr)
source('./r/stataknitrhelper.r')
opts_chunk$set(echo = FALSE, message = FALSE, comment = NA)
```

```{r, engine = 'bash'}
## run accompanying .do file to get log file for parsing
export PATH=$PATH:/Applications/Stata13/Stata/StataSE.app/Contents/MacOS
StataSE -e -q do ../do/lecture01_conditional_means.do
```

```{r}
## save log file in object
lf <- 'lecture01_conditional_means.log'
```

In almost all of our work, we seek to make predictions about what will 
happen under certain circumstances. Our basic tool for doing this is 
the conditional mean. Today we'll make the link between conditional 
means and regression.

# Unconditional Means

In most of our day-to-day thinking, we use unconditional means as the 
basis for making predictions. For instance, if I asked you to predict 
the temperature on June 1st of this year, you'd most likely simply say 
it would be the average temperature for that day or that time of 
year. 

In the example below, I calculate the mean of math test scores in the 
plans dataset and use it as a prediction. I calculate the error term,
then calculate the mean sqrm  uared error (which is exactly what it sounds 
like) as a measure of how good this prediction is. As the graphic 
shows, the mean is a pretty terrible predictor for most people.

It's worth remembering though, that the unconditional mean serves as
the basis for a very large amount of policymaking.


```{r}
start <- 'Using the mean as a prediction'
end <- 'Above average or median vs below average'
writeLines(logparse(lf, start = start, end = end))
```


![][../plots/uncond_mean.png]
# Conditonal Means

With condtional means, we start using more information to think about 
how we will make our prediction. One of the simplest ways to do this 
in a bivariate sense is to calculate the mean of the dependent 
variable for individuals who are above average and below average. 


```{r}
start <-'Above average or median vs below average' 
end <- 'Using quartiles as the basis for predictions'
writeLines(logparse(lf, start = start, end = end))
```

#Regression is the conditional mean

Regression is based on the idea of the expected value of y given,
E(Y|X). If X can take on only two values, then regression will give
two predictions. 

```{r}
start <- 'Using quartiles as the basis for predictions'
end <- 'End'
writeLines(logparse(lf, start = start, end = end))
```



